Finally Tagless, Partially Evaluated

### 2007-06-09 tagless/tagless-final.tex
Improved according to teh comments from the reviewers

\oleg{Emphasize somewhere in the Intro that we solved the problem
exactly as posed in Walid's, Xi's, and SPJ's papers. And we obtain the
result in the form those papers declare desirable.
We use their running example and achieve the result they
call desirable.}

@@

\oleg{We must stress that we do not Church-encode a universal type
(or the solution in the previous section). The latter will not be
patently total because we have to provide error continuatiuons and we
have to argue why they will not be called. A compiler is mots likely
incapable of such an argument and so has to compile them in. Here, we do
not have any error continuations at all.}

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is available%
\footnote{The URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

\oleg{We must stress that: of course all parts of our solution are
known. That was the whole point. The problem was claimed to be
unsolvable with the existing means, and so new techniques have been
proposed. We showed that regardless of the merits, power and convenience of
those new techniques, the existing, old solutions were sufficient for
the stated problem -- thus correcting the public record (of ICFP and
POPL conferences). We don't say GADT, etc. are useless -- but we wish
to show to what extent the existing old methods suffice.}
\oleg{Our contribution that we make do with a very restricted polymorphism
(HM plus the module system) which guarantees inference in ML and
Haskell.}
\oleg{So, our thrust is: there is public (ICFP, POPL) record of the problem
being open and claims that solving it requires various
extensions. We show that the problem as formulated in the literature
can be solved with the existing means. We claim the novelty of this,
essentially negative, result. We specifically disclaim any new
inventions.}

Again, the fact that our
solution is elementary and uses only existing known tools to give a
previously unknown solution to an important problem should not be
counted as a disadvantage.
We did not claim to invent a new method (of partial evaluation or of
interpretation). Rather, we claim that existing, old methods, when
properly combined, can solve problem that was publicly declared to be
unsolvable with the existing, old means.
Some parts of our solution have been known before; we mention that in
the paper and reviewers point more references. However, none of the
parts have been combined to solve the problem stated as open in
the recent published literature Walid03 and WalidTagless.

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent every typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

We have also implemented this compiler in Haskell. 
\oleg{mention TH solution in footnote. ``I was rather hoping it could
have been emulated via TemplateHaskell. I realise TH is untyped and a
compile-time staging system. The claim would have been that the TH
code produces will always be type correct. We could still map the
typed ByteCode data type to untyped TH abstract syntax.''}
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, but it may not; GHC for one
issues warnings.  Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The goal is to make this safety
\emph{syntactically} apparent.

\oleg{Rev3:In 4.3, please carefully distinguish the limitation of GHC's
implementation of GADTs from the type theory. The pattern match in 4.3
is statically exhaustive, and can be shown so in other systems. If
this code were implemented using inductive families Coq, for example,
the user would be able to (and in fact required) to prove that the
other cases could not occur.}

@@

Stress typed.

The new features may be deservingly
exciting and suitable. We do not wish to stop progress. We merely wish to set
the record straight as to which problem can be solved with which minimal and
widely available tools.

Even the idea of having multiple interpretations of the
pervasives of a standard language (as in the Symantics structure) has
been exploited by many (see e.g. the Lava and Hawk work, Augustsson's
clever two-level re-interpretation of Haskell (to generate residual
functions in Excel (!)), Rhiger's thesis, and many DSLs.

Stress differences from phantom types.
Use of phantom types is popular: see Lava 
Lava: Hardware Design in Haskell (1998)) and Rhiger's thesis.
(Lava: has universal type for Bit and NumSig and correspondent
tag matching: it is not tagless).

Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
OR:
The simplicity of our encoding enables trivial proofs of
soundness and completeness compared to Rhiger's.  It also makes
self-interpretation simple, which is a strength of our work.

first approach that gives typed CPS in a HM language. We also need to
cite the paper by Stefan Monnier at PLPV2006 where they say, I think,
that typed CPS is impossible without GADT or like.

the basic idea of this paper (the encoding of the term
GADT as its elimination form, i.e. as a coalgebraic structure) is not
new to this work. It appeared many years ago in: Metacircularity in
the polymorphic lambda-calculus, Frank Pfenning and Peter Lee.
Theoretical Computer Science 89 (1991) 137-159. (Other encodings of
GADTs have appeared since Pfenning and Lee's work, for example, in
work on generic programming.)
BTW,
many GADT encodings I've seen (Sulzmann's et al) require existentials,
which are not available in ML. The only GADT encodings I've seen in ML
use unsafe operations.

It seems that Pfenning and Lee embed $F_2$ with type constructions in 
(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
we do is very similar to what they do (Figure 1, p.152), except that we 
do it in standard programming languges.  It is unclear if their work can 
be implemented (yet) in any language.  And we preserve type-inference, 
while their solution needs explcit types!
The following line of their conclusion is worth citing: "... this does 
not imply that the same language is also suitable for type 
metaprogramming. ... such as partial evaluation... do not seem to be 
expressible".
I suspect you're right, but I'm still reading the paper.  See also page
146: "for a term M in $F_1$ (a simply-typed term), the representation
$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
and beyond reminds me
strongly of our attempts at self-interpretation without the notion of a
syntactic hole.
    Pfenning and Lee conclude that partial evaluation and program
    transformations "do not seem to be expressible" even using their
    extension to F-omega, perhaps because their avoidance of general
    recursive types compels them to include the polymorphic lift that we
    avoid in our Section 4.1.

 Although GADTs
may indeed be "more flexible and easier to use", they are not
available in mainstream ML, and GHC as released does not implement
them correctly (as Simon Peyton-Jones admits publicly).  Whereas
Pasalic et al. have already solved the problem of tagless typed
(staged) interpretation, our solution is the first to use only
facilities available in mainstream ML and Haskell, namely
Hindley-Milner and either simple modules or constructor classes.  The
simplicity of these facilities is again a strength of our work: it
makes embedded DSLs more practical to build.

Type-indexed types have been around since at least Harper and
Morrisett's Compiling Polymorphism Using Intensional Type Analysis,
POPL 1995. They also were supported by Generic Haskell (Hinze,
Juering, Loeh, Type Indexed Datatypes, SCP 2004).
Generic Haskell is not mainstream.  Xi et al. point out why
intensional type analysis cannot handle HOAS with fix.

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
embedded interpreters to justify
advanced type systems.  While we admire all this technical work,
we were not convinced that their 
motivating example actually needed all this machinery.

@@

Our final approach seems related to \citearound{'s approach to HOAS via
catamorphism and anamorphism}\citet{Washburn-Weirich-boxes}.

@@

@@

@@

\oleg{From Rev3}This representation limited to primitive
recursive folds over the datatype (as that is the elimination form
provided.) That means that some operations are easy implement
(interpretation, size calculation, etc.), some can be done with effort
(one-pass cps, and other mutually recursive functions), and perhaps
some cannot be implemented at all (those that are not primitive
recursive). For example, can you implement a transformation that
eliminates eta-expansions from a term?

>>>>

@@

Our solution is emphatically distinct from Church-encoding of the
universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
requires two continuations; the interpreter like |app| above would
have to provide both to the encoding of |e1|. The continuation
corresponding to the |UB| branch of |u| must either raise an error or
loop. For a well-typed object term that error continuation is never
invoked, yet it must be supplied. Our interpreter has no error
continuations at all.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is available%
\footnote{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

We take the problem of tagless (staged) typed-preserving
interpretation exactly as it was posed in \citet{WalidICFP02} and
\citet{xi-guarded}.  We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

We reiterate that we specifically do not propose any new language
feature or technique. We use the existing, known approaches --
Hindley-Milner type system with inference-preserving module system or
constructor classes extensions, as realized in ML or Haskell98 -- to
solve the problem that was stated in published record as lacking
solution in ML or Haskell98 and most likely unimplementable in these
languages without extensions. The simplicity of our solution and our
use of only mainstream functional language features makes typed
embedded DSLs more practical to build.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \citet{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

Comparing |Symantics| with Figure~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types, in the following
sense.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

@@

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.

@@

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the metalanguage implementation could reason thus, 
but it may not; GHC for one issues warnings%
\footnote{If we use inductive families (as those in Coq) or logical
  frameworks with canonical forms, we can prove the pattern-matching
  to be exhaustive. The coverage checker of Twelf does such proof.}. 
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The goal is to make this safety
\emph{syntactically} apparent.

@@

Our initial motivation came from several papers
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
embedded interpreters to justify advanced type systems, in particular,
GADT. While we admire all this technical work, we were not convinced
that their motivating example actually needed all this machinery.
Although GADTs may indeed be more flexible and easier to use, they are
not available in mainstream ML, and their current implementation in
GHC has problems. We were also interested in discovering the minimal and
widely available set of language features necessary for tagless
type-preserving interpretation.

@@

@@

The idea of multiple interpretations of the host language pervasives
such as addition or application is quite common in implementing
embedded DSL. It is also common to use phantom types to prevent
forming of ill-typed DSL terms \citet{Lava,Rhiger-thesis}. However,
the universal type still remains (|Bit| and |NumSig| in Lava
\citet{Lava}, |Raw| and |Term| in \citet[Fig 2.2,
  Sec. 3]{Rhiger-thesis}), along with the attendant pattern-matching
overhead. These approaches are not tagless.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\citet{Rhiger-thesis}. In contrast, our proofs become trivial.
The approach of \Citet[Sec 3.3.4]{Rhiger-thesis} does not 
support typed CPS transformation.
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}

@@

Our approach is based on encoding a term in its elimination form, as a
co-algebraic structure. That basic idea and its application to
meta-circular interpertation first appeared in \citet{Pfenning-Lee}.
Our approach however is implementable in mainstream ML and supports
type inference and typed CPS and partial evaluation. In contrast,
\citet{Pfenning-Lee} conclude that partial evaluation and program
transformations "do not seem to be expressible" even using their
extension to $F_{\omega}$, perhaps because their avoidance of general
recursive types compels them to include the polymorphic lift that we
avoid in \S\ref{S:PE-lift}.
\begin{comment}
It seems that Pfenning and Lee embed $F_2$ with type constructions in 
(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
we do is very similar to what they do (Figure 1, p.152), except that we 
do it in standard programming languges.  It is unclear if their work can 
be implemented (yet) in any language.  And we preserve type-inference, 
while their solution needs explcit types!
The following line of their conclusion is worth citing: "... this does 
not imply that the same language is also suitable for type 
metaprogramming. ... such as partial evaluation... do not seem to be 
expressible".
I suspect you're right, but I'm still reading the paper.  See also page
146: "for a term M in $F_1$ (a simply-typed term), the representation
$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
and beyond reminds me
strongly of our attempts at self-interpretation without the notion of a
syntactic hole.
\end{comment}

Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
may be regarded as an emulation of type-indexed types and related to
the intensional type analysis
\citet{Morrisett-intensional,Generic-Haskell}. Our running HOAS
example however, includes |fix|, which cannot be handled by the
intensional type analysis \citet{xi-guarded}.  Our final approach
seems related to \citearound{'s approach to HOAS via catamorphism and
  anamorphism}\citet{Washburn-Weirich-boxes}.

@@

Our representation of object terms in the elimination form encodes
primitive recursive folds over the terms. This makes operations like
interpretation and size calculations trivial to implement. It remains
the subject of the current work to see if and how non-primitively 
recursive operations can be supported in our approach.
